---
title: "Simulating and Plotting Counterfactuals"
author: "David Johnston"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(here::here(), "docs", "create_report.html")) })
output: html_document
---
## R Packages

```{r, echo=TRUE, eval=FALSE}
# MRC-IDE Packages
devtools::install_github(
  "mrc-ide/squire.page",
  ref = "3fbb97f3e72c4e6bd7f66690a53f7bb59ce5ea0e"
) # should install odin, squire and nimue
devtools::install_github(
  "mrc-ide/drjacoby",
  ref = "476d94f3eb7357f8e2278834c0af04afd772cf69"
)
# CRAN packages
install.packages("orderly")
install.packages("countrycode") # version 1.3.1
install.packages("future") # version 1.24.0
install.packages("dplyr") # version 1.0.8
install.packages("stringr") # version 1.4.0
install.packages("purrr") # version 0.3.4
install.packages("tidyr") # version 1.2.0
install.packages("readr") # version 2.1.2
install.packages("lubridate") # version 1.8.0
install.packages("tibble") # version 3.1.6
install.packages("rlang") # version 1.0.2
install.packages("ggplot2") # version 3.3.5
install.packages("forcats") # version 0.5.1
install.packages("cowplot") # version 1.1.1
install.packages("ggpubr") # version 0.4.0
install.packages("gridExtra") # version 2.3
install.packages("sf") # version 1.0.7
install.packages("qpdf") # version 1.1
install.packages("knitr")
install.packages("here")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
```



All packages used in this repository and the versions used to generate the data.
Feel free to skip if you already have these.

## Running the simulation

The following simulates the baseline pandemic in the United States, as well as a
collection of predefined counterfactuals. If you are looking to alter the counterfactuals
simulated, you can edit this file.

```{r, echo=TRUE, eval=FALSE}
source("data/raw/generate_counterfactuals.R")
```

## Running orderly reports

### Set the directory

```{r, echo=TRUE, eval=FALSE}
setwd(root_path)
orderly_loc <- here::here()
```

### Get Model Fits

The model fits used to generated the data in this repository are stored in the 
*data* directory. The following tasks gather the simulated counterfactuals from
this directory.

```{r, echo=TRUE, eval=FALSE}
gather_simulations_time_id <- orderly::orderly_run("gather_simulations_time",
  parameters = list(
    orderly_loc = orderly_loc,
    excess = TRUE
  ), echo = FALSE
)
orderly::orderly_commit(gather_simulations_time_id)
```

### Create the plots

```{r, echo=TRUE, eval=FALSE}
d_a_plot_id <- orderly::orderly_run("deaths_averted_plot_timeline",
  parameters = list(excess = TRUE)
)
orderly::orderly_commit(d_a_plot_id)
```

