---
title: "Simulating and Plotting Counterfactuals"
author: "David Johnston"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(here::here(), "docs", "create_report.html")) })
output: html_document
---
## R Packages

```{r, echo=TRUE, eval=FALSE}
# MRC-IDE Packages
devtools::install_github(
  "mrc-ide/squire.page",
  ref = "fc27dcabec55d70ec293328813cec644c56a63db"
) # should install odin, squire and nimue
devtools::install_github(
  "mrc-ide/drjacoby",
  ref = "476d94f3eb7357f8e2278834c0af04afd772cf69"
)
# CRAN packages
install.packages("orderly")
install.packages("countrycode") # version 1.3.1
install.packages("future") # version 1.24.0
install.packages("dplyr") # version 1.0.8
install.packages("stringr") # version 1.4.0
install.packages("purrr") # version 0.3.4
install.packages("tidyr") # version 1.2.0
install.packages("readr") # version 2.1.2
install.packages("lubridate") # version 1.8.0
install.packages("tibble") # version 3.1.6
install.packages("rlang") # version 1.0.2
install.packages("ggplot2") # version 3.3.5
install.packages("forcats") # version 0.5.1
install.packages("cowplot") # version 1.1.1
install.packages("ggpubr") # version 0.4.0
install.packages("gridExtra") # version 2.3
install.packages("sf") # version 1.0.7
install.packages("qpdf") # version 1.1
install.packages("knitr")
install.packages("here")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
```



All packages used in this repository and the versions used to generate the data.
Feel free to skip if you already have these.

## Running orderly reports

### Set the directory and parameters

```{r, echo=TRUE, eval=FALSE}
setwd(root_path)
date <- '2023-01-01'
```

### Run the fits if necessary

Note that running the model fits takes a long time and downloads an 8GB sequence file every time you do it. Run only if necessary.

The following runs preliminary tasks for generating the model fits. It does not take too long, but `input_sequencing` downloads an 8GB sequence file every time it's run.

```{r, echo=TRUE, eval=FALSE}
cm <- orderly::orderly_run("parameters_vaccines", echo = FALSE)
orderly::orderly_commit(cm)

rt_ord_runs <- lapply(
  list(
    "input_excess_mortality",
    "input_jhu",
    "input_vaccinations"
  ),
  function(x) {
    cm <- orderly::orderly_run(x, parameters = list(date = date), echo = FALSE)
    orderly::orderly_commit(cm)
  }
)

seq_id <- orderly::orderly_run("input_sequencing", parameters = list(date = date, gisaid = FALSE), echo=FALSE)
orderly::orderly_commit(seq_id)

```

The following runs the model fits for USA and UK. It taks 3-9 hours on my m1 macbook.

```{r, echo=TRUE, eval=FALSE}
iso3cs <- c("USA", "GBR")

rt_params_lists <- lapply(iso3cs, function(iso3c){
  list(
      iso3c = iso3c,
      date = date,
      samples = 32*3, #how many random samples to generate and fit
      seed = FALSE, #Set a seed, useful for debugging
      parallel = TRUE,
      #Should we build the required documentation+data for the fit
      document = TRUE,
      fit_cases = FALSE #issues with this, don't have time to fix them.
      # #Fitting parameters (leave as blank to use defaults in fitting_params.Rds)
      # initial_infections_interval = c(5, 500),
      # n_particles = 10,
      # k = 14,
      # rt_interval = c(0.5, 10)
    )
})

for (rt_params in rt_params_lists) {
  rt_id <- orderly::orderly_run(
    "lmic_reports_rt_optimise",
    parameters = rt_params,
    echo = FALSE)

  orderly::orderly_commit(rt_id)
}
```


### Run the simulations

The following simulates the baseline pandemic in the United States and the United Kingdom, as well as a
collection of predefined counterfactuals. If you are looking to alter the counterfactuals
simulated, you can edit this file.

```{r, echo=TRUE, eval=FALSE}
get_baseline_production_id <- orderly::orderly_run("get_baseline_production",
                                                   echo = FALSE 
)
orderly::orderly_commit(get_baseline_production_id)
```

```{r, echo=TRUE, eval=FALSE}
generate_production_counterfactual_id <- orderly::orderly_run("generate_production_counterfactual",
                                                          parameters = list(
                                                              shift_by = 60,
                                                              interpolation_source_end = '2022-01-01'
                                                          ), echo = FALSE 
)
orderly::orderly_commit(generate_production_counterfactual_id)
```

```{r, echo=TRUE, eval=FALSE}
generate_vaccine_counterfactual_id <- orderly::orderly_run("generate_vaccine_counterfactual",
                                                          parameters = list(
                                                              shift_by = 60
                                                          ), echo = FALSE
)
orderly::orderly_commit(generate_vaccine_counterfactual_id)
```

There are two counterfactual models available; a newer one with booster doses and explicit modelling of omicron and an older one with neither of these. The older models results differ somewhat from the newer model's results, so it may be of interest in understanding what the results are sensitive to.

For the new model, to generate a counterfacutual simulation, run the following. The parameter `excess` controls whether the baseline course of infection was fit to confirmed cases or excess deaths. The parameter `lowtransmission` controls whether the relative risk of onward transmission for vaccinated individuals is set to 0.5, for consistency with the older model, or to 0.78 for consistency with published data. Unless you're investigating the impact of this parameter, you should set it to false.

```{r, echo=TRUE, eval=FALSE}
generate_counterfactuals_id <- orderly::orderly_run("generate_counterfactuals",
                                                   parameters = list(
                                                        excess = TRUE,
                                                        lowtransmission = FALSE
                                                   ), echo = FALSE
)
orderly::orderly_commit(generate_counterfactuals_id)
```

To run the older model, use the following command instead:

```{r, echo=TRUE, eval=FALSE}
generate_counterfactuals_old_id <- orderly::orderly_run("generate_counterfactuals_oldfits",
                                                   parameters = list(
                                                        excess = TRUE
                                                   ), echo = FALSE
)
orderly::orderly_commit(generate_counterfactuals_old_id)
```

### Create the plots

Plots for the counterfactual simulations can be generated with the following task. Here, the parameters `excess` and `lowtransmission` determine which counterfactual simulation is plotted. The task will use the most recent counterfactual simulation for which the values of both parameters matches; if there is no simulation that matches both parameters, it will fail.

```{r, echo=TRUE, eval=FALSE}
d_a_plot_id <- orderly::orderly_run("deaths_averted_plot_timeline",
  parameters = list(
      excess = TRUE,
      lowtransmission = FALSE
  )
)
orderly::orderly_commit(d_a_plot_id)
```

To create plots for the older simulation, use

```{r, echo=TRUE, eval=FALSE}
d_a_plot_old_id <- orderly::orderly_run("deaths_averted_plot_timeline_oldfits",
  parameters = list(
      excess = TRUE,
  )
)
orderly::orderly_commit(d_a_plot_old_id)
```