---
title: "Simulating and Plotting Counterfactuals"
author: "David Johnston"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
knit: (function(inputFile, encoding) { 
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(here::here(), "docs", "create_report.html")) })
output: html_document
---
## R Packages

```{r, echo=TRUE, eval=FALSE}
# MRC-IDE Packages
devtools::install_github(
  "mrc-ide/squire.page",
  ref = "fc27dcabec55d70ec293328813cec644c56a63db"
) # should install odin, squire and nimue
devtools::install_github(
  "mrc-ide/drjacoby",
  ref = "476d94f3eb7357f8e2278834c0af04afd772cf69"
)
# CRAN packages
install.packages("orderly")
install.packages("countrycode") # version 1.3.1
install.packages("future") # version 1.24.0
install.packages("dplyr") # version 1.0.8
install.packages("stringr") # version 1.4.0
install.packages("purrr") # version 0.3.4
install.packages("tidyr") # version 1.2.0
install.packages("readr") # version 2.1.2
install.packages("lubridate") # version 1.8.0
install.packages("tibble") # version 3.1.6
install.packages("rlang") # version 1.0.2
install.packages("ggplot2") # version 3.3.5
install.packages("forcats") # version 0.5.1
install.packages("cowplot") # version 1.1.1
install.packages("ggpubr") # version 0.4.0
install.packages("gridExtra") # version 2.3
install.packages("sf") # version 1.0.7
install.packages("qpdf") # version 1.1
install.packages("knitr")
install.packages("here")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
```



All packages used in this repository and the versions used to generate the data.
Feel free to skip if you already have these.

## Running orderly reports

### Set the directory and parameters

```{r, echo=TRUE, eval=FALSE}
setwd(root_path)
date <- '2023-01-01'
```

### Run the fits if necessary

Note that running the model fits takes a long time and downloads an 8GB sequence file every time you do it. Run only if necessary.

The following runs preliminary tasks for generating the model fits. It does not take too long, but `input_sequencing` downloads an 8GB sequence file every time it's run.

```{r, echo=TRUE, eval=FALSE}
cm <- orderly::orderly_run("parameters_vaccines", echo = FALSE)
orderly::orderly_commit(cm)

rt_ord_runs <- lapply(
  list(
    "input_excess_mortality",
    "input_jhu",
    "input_vaccinations"
  ),
  function(x) {
    cm <- orderly::orderly_run(x, parameters = list(date = date), echo = FALSE)
    orderly::orderly_commit(cm)
  }
)

seq_id <- orderly::orderly_run("input_sequencing", parameters = list(date = date, gisaid = FALSE), echo=FALSE)
orderly::orderly_commit(seq_id)

```

The following runs the model fits for USA and UK. It taks 3-9 hours on my m1 macbook.

```{r, echo=TRUE, eval=FALSE}
iso3cs <- c("USA", "GBR")

rt_params_lists <- lapply(iso3cs, function(iso3c){
  list(
      iso3c = iso3c,
      date = date,
      samples = 32*3, #how many random samples to generate and fit
      seed = FALSE, #Set a seed, useful for debugging
      parallel = TRUE,
      #Should we build the required documentation+data for the fit
      document = TRUE,
      fit_cases = FALSE #issues with this, don't have time to fix them.
      # #Fitting parameters (leave as blank to use defaults in fitting_params.Rds)
      # initial_infections_interval = c(5, 500),
      # n_particles = 10,
      # k = 14,
      # rt_interval = c(0.5, 10)
    )
})

for (rt_params in rt_params_lists) {
  rt_id <- orderly::orderly_run(
    "lmic_reports_rt_optimise",
    parameters = rt_params,
    echo = FALSE)

  orderly::orderly_commit(rt_id)
}
```


### Run the simulations

The following simulates the baseline pandemic in the United States and the United Kingdom, as well as a
collection of predefined counterfactuals. If you are looking to alter the counterfactuals
simulated, you can edit this file.

```{r, echo=TRUE, eval=FALSE}
generate_production_counterfactual_id <- orderly::orderly_run("generate_production_counterfactual",
                                                          parameters = list(
                                                              shift_by = 60,
                                                              interpolation_source_end = '2022-01-01'
                                                          ), echo = FALSE 
)
orderly::orderly_commit(generate_production_counterfactual_id)
```

```{r, echo=TRUE, eval=FALSE}
generate_vaccine_counterfactual_id <- orderly::orderly_run("generate_vaccine_counterfactual",
                                                          parameters = list(
                                                              shift_by = 60
                                                          ), echo = FALSE
)
orderly::orderly_commit(generate_vaccine_counterfactual_id)
```

There are two counterfactual models available; a newer one with booster doses and explicit modelling of omicron and an older one with neither of these. The older models results differ somewhat from the newer model's results, so it may be of interest in understanding what the results are sensitive to.

For the new model, to generate a counterfacutual simulation, run the following. The parameter `excess` controls whether the baseline course of infection was fit to confirmed cases or excess deaths, `boosters` controls whether booster doses are included in the simulation or set to 0 and `double_boosters` runs a counterfactual where the rate of booster uptake for the US is similar to the uptake in the UK. We run the simulation with three different parameter sets in order to generate the report.

```{r, echo=TRUE, eval=FALSE}

tasks <- list(
  "generate_counterfactuals",
  "deaths_averted_plot_timeline")

parameter_sets <- list(
    list(
      excess=FALSE, 
      boosters=TRUE,
      double_boosters = FALSE
    ),
    list(
      excess=TRUE, 
      boosters=TRUE,
      double_boosters = FALSE
    ),
    list(
      excess=TRUE, 
      boosters=TRUE,
      double_boosters = TRUE
    ) 
  )

ord_runs <- lapply(
  tasks,
  function(task) {
    lapply(
      parameter_sets,
      function(params) {
        cm <- orderly::orderly_run(task, parameters = params, echo = FALSE)
        orderly::orderly_commit(cm)
      })
  }
)
```

To run the older model, use the following command instead. Note that the older model is not integrated with the preprint pipeline.

```{r, echo=TRUE, eval=FALSE}
generate_counterfactuals_old_id <- orderly::orderly_run("generate_counterfactuals_oldfits",
                                                   parameters = list(
                                                        excess = TRUE
                                                   ), echo = FALSE
)
orderly::orderly_commit(generate_counterfactuals_old_id)
```

### Generate the report

To generate the report, run the following

```{r, echo=TRUE, eval=FALSE}
preprint <- orderly::orderly_run("preprint", echo = FALSE)
orderly::orderly_commit(preprint)
```

For the older simulation, plots can be generated with

```{r, echo=TRUE, eval=FALSE}
d_a_plot_old_id <- orderly::orderly_run("deaths_averted_plot_timeline_oldfits",
  parameters = list(
      excess = TRUE,
  )
)
orderly::orderly_commit(d_a_plot_old_id)
```