# Results

## Vaccinations under counterfactual scenarios

Figure \@ref(fig:cumul-vacc-cfacts) shows the cumulative vaccinations for the counterfactual scenarios we investigate. The baseline scenario is the actual vaccination timeline, while the other scenarios are the same timeline but with vaccines administered 30, 60 or 90 days sooner.

```{r cumul-vacc-cfacts, fig.align="center", fig.cap=c("DJ4 Cumulative vaccine counterfactuals"), echo=FALSE}
baseline = readRDS("data/owid_raw.Rds")
sooner_30 = readRDS("data/30_days_sooner.Rds")
sooner_60 = readRDS("data/60_days_sooner.Rds")
sooner_90 = readRDS("data/90_days_sooner.Rds")

counterfactuals = bind_rows(baseline, sooner_30, sooner_60, sooner_90) %>%
  filter(date <= primary_last_date)

dj4 = ggplot() +
  geom_line(data = counterfactuals,
            aes(x = as.Date(date),
                y = total_vacc,
                color = shifted_by)) +
  facet_wrap(~country, ncol = 2, scales = "free_y") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  scale_colour_manual(
    labels = c('30 days', '60 days', '90 days', 'baseline'),
    values = c(alpha(colour_counterfactual, 0.33), alpha(colour_counterfactual, 0.66), colour_counterfactual, colour_baseline)
  ) +
  labs(x = "Date",
       y = "Cumulative vaccinations",
       color = "Vaccinations starting sooner by:")

plot(dj4)
```

Table \@ref(tab:vaccinations-table) shows the actual vaccinations administered on 2021-01-01, 2021-04-01 and 2021-07-01 alongside the administration numbers for the counterfactual scenarios we investigate.


```{r vaccinations-table, caption=c("Vaccinations administered under baseline and counterfactual scenarios"), echo=FALSE}
vaccinations_table = counterfactuals %>%
  filter(date %in% c(
    as.Date('2021-01-01'),
    as.Date('2021-04-01'),
    as.Date('2021-07-01')
  )) %>%
  select(country, date, total_vacc, shifted_by) %>%
  mutate(
    total_vacc = format(total_vacc,
                        big.mark = ",",
                        trim = TRUE),
    shifted_by = mapvalues(
      shifted_by,
      from = c("baseline", "30", "60", "90"),
      to = c(
        "Baseline",
        "Vaccines 30 days sooner",
        "Vaccines 60 days sooner",
        "Vaccines 90 days sooner"
      )
    )
  ) %>%
  plyr::rename(
    c(
      "shifted_by" = "Counterfactual scenario"
    )
  ) %>%
  pivot_wider(names_from = date,
              values_from = total_vacc) %>%
  replace(is.na(.), '0') %>%
  arrange(country) %>%
  select(-country)

vaccinations_table %>%
  kbl(booktabs=TRUE, centering = TRUE, caption = "Total vaccinations at sample dates") %>%
  pack_rows("United Kingdom", 1, 4) %>%
  pack_rows("United States", 5, 8)

```

## Impact of counterfactual vaccination scenarios

Table \@ref(tab:deaths-averted-table) shows the number of deaths averted by each counterfactual scenario. The number of deaths averted is calculated as the difference between the number of deaths under the baseline scenario and the number of deaths under the counterfactual scenario. The number of deaths averted is shown as the average number of deaths averted, as well as the interval that contains 95% of simulation trajectories (see Section [Methods] for details on how simulation trajectories are sampled). The number of deaths averted is also shown as the average number of deaths averted per 10,000 people in each country.

```{r deaths-averted-table, echo=FALSE}

differences_table <- deaths_averted_table_single(load_deaths_averted_table())

differences_table %>%
  kbl(booktabs=TRUE, centering = TRUE, caption = "DJ5 Averted deaths") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)

```


Figure \@ref(fig:deaths-averted-plot) shows the number of deaths per day under each counterfactual scenario compared to the baseline. Also shown is the interval containing 95% of the counterfactual simulation trajectories. Figure \@ref{cumulative-deaths} shows cumulative deaths under the baseline and counterfactual scenarios.

```{r deaths-averted-plot, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario"), echo=FALSE}
deaths_averted = readRDS("data/deaths_averted_detail.Rds")

zoomed_deaths_averted = deaths_averted %>%
  filter(date > as.Date("2020-09-01") & date <= primary_last_date) %>%
  filter(counterfactual %in% cfacts_to_display) %>%
  mutate(counterfactual_label = mapvalues(
    counterfactual,
    from = cfacts_to_display,
    to = labels_to_display
  )) %>%
  mutate(counterfactual_label = factor(counterfactual_label,
                                       labels_to_display)) %>%
  left_join(population, by = "iso3c") %>%
  mutate(
    deaths_025_perpop = deaths_025 / pop10k,
    deaths_975_perpop = deaths_975 / pop10k,
    baseline_deaths_avg_perpop = baseline_deaths_avg / pop10k,
    deaths_avg_perpop = deaths_avg / pop10k,
    cumulative_deaths_025_perpop = cumulative_deaths_025 / pop10k,
    cumulative_deaths_975_perpop = cumulative_deaths_975 / pop10k,
    baseline_cumulative_deaths_avg_perpop = baseline_cumulative_deaths_avg / pop10k,
    cumulative_deaths_avg_perpop = cumulative_deaths_avg / pop10k,
  )
  
baseline_color <- "#2C3E50" # dark blue
counterfactual1_color <- "#E74C3C" # red
counterfactual2_color <- "#27AE60" # green
counterfactual3_color <- "#F1C40F" # yellow

single_timeseries_plot = ggplot(data = zoomed_deaths_averted,
                                aes(x = as.Date(date),
                                    y = deaths_avg_perpop,
                                    color = counterfactual)) +
  geom_line() +
  geom_ribbon(
    aes(ymin = deaths_025_perpop, ymax = deaths_975_perpop, fill = counterfactual),
    alpha = 0.2,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(aes(y = baseline_deaths_avg_perpop, colour = "baseline")) +
  facet_wrap( ~ country) +
  scale_colour_manual(
    aesthetics = c("fill", "colour"),
    labels = c('30 days', '60 days', '90 days', 'baseline'),
    values = c(
      counterfactual1_color,
      counterfactual2_color,
      counterfactual3_color,
      baseline_color
    )
  ) +
  labs(x = "Date", y = "Daily Deaths per 10 000 people", color = "Earlier vaccines by:")

plot(single_timeseries_plot)


```


```{r cumulative-deaths, fig.align="center", fig.cap=c("Cumulative deaths per scenario"), echo=FALSE}
cumulative_deaths_plot = ggplot(zoomed_deaths_averted,
                                aes(x = as.Date(date),
                                    y = cumulative_deaths_avg_perpop,
                                    color = counterfactual)) +
  geom_line() +
  geom_ribbon(
    aes(ymin = cumulative_deaths_025_perpop, ymax = cumulative_deaths_975_perpop, fill = counterfactual),
    alpha = 0.2,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(aes(y = baseline_cumulative_deaths_avg_perpop, colour = "baseline")) +
  facet_wrap( ~ country) +
  scale_colour_manual(
    aesthetics = c("fill", "colour"),
    labels = c('30 days', '60 days', '90 days', 'baseline'),
    values = c(
      counterfactual1_color,
      counterfactual2_color,
      counterfactual3_color,
      baseline_color
    )
  ) +
  labs(x = "Date", y = "Cumulative Deaths per 10 000 people", color = "Earlier vaccines by:")
  

plot(cumulative_deaths_plot)
```


