BIBLIOGRAPHY


# (APPENDIX) Appendix {-}

# Sensitivity

## Baseline fit to reported deaths instead of excess deaths

```{r reported-deaths-averted-table, echo=FALSE}

differences_table <- deaths_averted_table_single(load_deaths_averted_table( excess = FALSE ))

differences_table %>%
  kbl(booktabs=TRUE, centering = TRUE, caption = "Averted deaths calculated on the basis of a model fit to reported deaths") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)

```

```{r reported-deaths-averted-plot, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario"), echo=FALSE}
deaths_averted = readRDS("data/reported_deaths_averted_detail.Rds")

zoomed_deaths_averted = deaths_averted %>%
  filter(date > as.Date("2020-09-01") & date <= primary_last_date) %>%
  filter(counterfactual %in% cfacts_to_display) %>%
  mutate(counterfactual_label = mapvalues(
    counterfactual,
    from = cfacts_to_display,
    to = labels_to_display
  )) %>%
  mutate(counterfactual_label = factor(counterfactual_label,
                                       labels_to_display))

single_timeseries_plot = ggplot(zoomed_deaths_averted, aes(x = date)) +
  geom_line(aes(y = baseline_deaths_avg, colour = "baseline")) +
  geom_line(aes(y = deaths_avg, colour = "counterfactual")) +
  geom_ribbon(
    aes(ymin = deaths_025, ymax = deaths_975),
    alpha = 0.3,
    fill = colour_counterfactual
  ) +
  facet_grid(counterfactual_label ~ country,
             labeller = labeller(.rows=label_wrap_gen(width = 15))) +
  scale_colour_manual(
    labels = c('baseline', 'counterfactual'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  labs(x = "Date", y = "Daily Deaths", color = "Colour")
  

plot(single_timeseries_plot)


```

## Counterfactual scenario: what if US had doubled the rate of booster uptake?

```{r deaths-averted-table-doubleboost, echo=FALSE, results='asis'}
differences_table <- deaths_averted_table_single(
    load_deaths_averted_table(
      table_path = "data/doubleboost_deaths_averted_detail.Rds")
  )

differences_table %>% kbl(booktabs = TRUE, centering = TRUE, caption = "How would US outcomes change if boosters were adopted at twice the actual rate?") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)


```

## Sensitivity to parametric estimates

```{r deaths-averted-table-vei, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))

differences_tables$veis %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to average vaccine efficacy against infection (VEI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)


```

```{r deaths-averted-table-durV, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))

differences_tables$dur_Vs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of vaccine acquired immunity (DVI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)

```

```{r deaths-averted-table-durR, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))

differences_tables$dur_Rs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of naturally acquired immunity (DNI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)

```

```{r deaths-averted-plot-vei, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to vaccine duration"), echo=FALSE}

deaths_averted_plots <- deaths_averted_plot('dur_Vs', percentiles, last_dates_to_display)

```

```{r deaths-averted-plot-durV, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to vaccine efficacy"), echo=FALSE}

deaths_averted_plots <- deaths_averted_plot('veis', percentiles, last_dates_to_display)

```

```{r deaths-averted-plot-durR, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to natural immunity duration"), echo=FALSE}

deaths_averted_plots <- deaths_averted_plot('dur_Rs', percentiles, last_dates_to_display)
```

```{r vaccinated-plot-vei, fig.align="center", fig.cap=c("Effective vaccinated, sensitivity to vaccine duration"), echo=FALSE}

deaths_averted_plots <- vaccinated_plot('dur_Vs', percentiles, last_dates_to_display[3])

```

# Vaccine production model

Figure \@ref(fig:prod-assumptions) shows the assumptions we made about the achievable production when vaccines were approved to be used 30, 60 and 90 days earlier.

```{r prod-assumptions, fig.align="center", fig.cap=c("DJ9 Vaccine production assumptions"), echo=FALSE}
detail_prod = readRDS("data/detail_production.Rds")
pfizer_prod = readRDS("data/pfizer_estimated_production.Rds")
moderna_prod = readRDS("data/moderna_estimated_production.Rds")

plot1 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = cumulative_available_vaccines)) +
  xlab("Date") +
  ylab("Cumulative vaccines available") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot2 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = daily_vaccines)) +
  xlab("Date") +
  ylab("Daily vaccines produced") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot3 = ggplot() +
  geom_line(data = moderna_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nModerna") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot4 = ggplot() +
  geom_line(data = pfizer_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nPfizer") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

combined_plot =
  ggarrange(plot3,
            plot4,
            plot2,
            plot1,
            nrow = 2,
            ncol = 2)
plot(combined_plot)
```


Figures \@ref(fig:prod-cap-uk) and \@rel{fig:prod-cap-usa} show our assumptions about the maximum number of vaccines that could be produced each day if vaccines were approved 30, 60 and 90 days earlier.

```{r prod-cap-uk, fig.align="center", fig.cap=c("DJ10 Limits from production UK"), echo=FALSE}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United Kingdom"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)
```


```{r prod-cap-usa, fig.align="center", fig.cap=c("DJ10 Limits from production UK"), echo=FALSE}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United States"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)

```