BIBLIOGRAPHY

---
output:
  pdf_document: default
  html_document: default
---

# Appendix A - Vaccine manufacturing in 2020

## Manufacturing constraints

If vaccines were approved earlier, production capacity would probably have been a major constraint on the number of doses delivered at the beginning of the mass vaccination campaigns. As mentioned in the main text, hypothesising the extent to which production is a binding constraint is complicated by the fact that new manufacturing capacity (or expansion of existing capacity) was put in place in anticipation of regulatory approval. Therefore it is likely that an earlier approval would have led the vaccine makers to expand production faster. In our base case model we ignore this, making our assumption on available doses a conservative one. 

While we model the United States and the United Kingdom using separate epidemiological models, we do not differentiate between countries when it comes to manufacturing capacity, reasoning that if a country was able to approve a vaccine earlier it would jump to the front of the vaccine deliveries queue. We also ignore Astrazeneca vaccine production since the production of Pfizer and Moderna vaccines are enough to satisfy demand in the UK in the initial phase of vaccinations. (As a conservative assumption, we still assume that vaccine effectiveness in the UK is a mix of mRNA vaccines and the one manufactured by AstraZeneca.)

To estimate the potential number of vaccines that would have been available throughout 2020, we used information about the actual manufacturing timelines. We assume that in 2020 Moderna produced 20 million doses[^21] and Pfizer/BioNTech produced 50 million[^22]. Additionally, we know Moderna was producing 500,000 doses a day at the time their vaccine was authorised[^23] and assume Pfizer/BioNTech was producing 1 million doses a day at the time they received authorization[^24]. These numbers, combined with estimates of scale-up dates, give us a possible function of daily vaccine production over 2020.

We assume a linear scale-up of production from March to July, followed by constant production until two weeks before authorization by the FDA. Specifically, we assume Pfizer/BioNTech started working on scale up from their announced partnership on March 17th[^25] and Moderna from March 23rd[^26]. The start of mass production in July (for calculations we use July 31st) was assumed based on reports that Pfizer might have been ready to mass produce by then[^27], with Moderna entering a partnership with Catalent in June to do their fill-and-finish processing with an estimated start at the beginning of Q3[^28]. Time of vaccine authorization was December 11th Pfizer/BioNTech[^29], and December 18th for Moderna[^30] meaning another scale up in manufacturing started on November 27th for Pfizer and December 4th for Moderna.

At this point, we assume companies already knew that they would most likely get authorization[^31] and were able to do another linear scale-up to a production level they would maintain until the end of the year. After the end of 2020 we assume production is unconstrained, i.e. it is as if infinite vaccines were available each day since 01-01-2021. This assumption is not important since even with a 90 day shift of vaccination series the production constraint stops being binding before the end of 2020. Additionally, we assume that if a country did approve vaccines earlier, they would nearly have all manufactured vaccines to themself. 

In the case of the United Kingdom this means that in all scenarios of earlier vaccination the production constraint does not bind at all, since the amount of vaccines is very large compared with their population size. In the case of the United States the production only binds in the case when the vaccination is shifted by 90 days.

## Vaccine hesitancy

A useful framework for vaccine hesitancy is provided by the “5C’s” model, which considers the following drivers: confidence in vaccine safety, complacency, convenience, risk calculation, and sense of collective responsibility [@Machingaidze2021Understanding]. For the purposes of our model, confidence in vaccine safety is the factor most clearly affected by a hypothetical earlier approval. We used surveys evaluating intention to be vaccinated and surveys evaluating confidence in health authorities to approximate the effect of vaccine hesitancy within our timeframe of interest. 

In the US and UK, the rate of respondents who intended to be vaccinated when vaccines were approved was 41% and 73% respectively. These proportions at 90 days prior to approval were 39% in the US and 71% in the UK, with the lowest rates of intention to be vaccinated at 36% in October 2020 in the US and 61% in November 2020 in the UK [@YouGov2023COVID, @Freeman2022COVID]. In a separate series of surveys in the US, 85% of respondents had a lot or a fair amount of confidence in health authorities at time of vaccine approval compared to 89% in April 2020 [@Kennedy2022Americans]. In the UK, surveys asking respondents their confidence in the country’s health system indicated 73% were confident at time of approval and 77% were confident 90 days prior [@YouGov2023Confidence]. By these estimations, vaccine hesitancy and confidence in health authorities hovered around similar levels during our period of interest. 

Dzieciolowska et al. conducted a study of 2,761 health care workers in Canada in December 2020 to assess vaccine hesitancy, with 19.1% of participants indicating they were not willing to receive the COVID vaccine[@Dzieciolowska2021COVID]. Among participants who refused the vaccine, 82% indicated the novelty of the vaccine was very or somewhat important in their decision and 60% indicated that they did not have enough time to make a decision. An approval of vaccines using accelerated timelines could obviously negatively impact hesitancy rooted in vaccine novelty and feelings of needing more time to decide . However, the main hypothetical we discuss in our paper is accelerating efficacy testing and starting safety testing early. This also suggests that while the impact may be non-negligible, it may not be large.

Joshi et al. found that the global average vaccine acceptance rate was 86% in March 2020, dropped to 54% in July 2020, and increased to 72% in September 2020[@Joshi2021Predictors].  By these estimates, if vaccines were approved during summer of 2020, hesitancy may have appreciably impacted uptake. However, it appears that vaccine hesitancy would likely not have impacted uptake if vaccines had been available up to 90 days earlier in the UK and US, with surveys evaluating intention to be vaccinated and trust in health authorities varying by no more than a few percentage points during this time. It is possible these trends would have been impacted by a concern among the general public for vaccines approved quickly, but these effects are difficult to quantify.

# Appendix B - Supplementary methods

## Baseline model

To model alternative vaccination timelines, we took an existing compartmental model that simulated the COVID-19 pandemic through 2020 and 2021 including the impact of vaccinations (the "baseline model") and re-ran the simulation freezing all parameters except the vaccination timelines (the "counterfactual models").

The baseline pandemic model used the model behind [@watsonoliverj.COVID19LMICReports2022]. This model is implemented in open source R packages Squire [@hoganWithincountryAgebasedPrioritisation2021, @walkerImpactCOVID19Strategies2020, @watsonLeveragingCommunityMortality2021] and SirCOVID [@baguelinSircovidSIRModel2022]. An older version of this model is described in detail in the supplementary material of [@walkerImpactCOVID19Strategies2020], and details of many parameters -- again, for an older version of the model -- can be found in the supplementary material of [@watsonGlobalImpactFirst2022]. Additional incomplete information about the parametrisation of the version of the model we used can be found at [COVID-19 LMIC Reports documentation](https://web.archive.org/web/20221230232942/https://mrc-ide.github.io/global-lmic-reports/parameters.html) [@watsonoliverj.COVID19LMICReports2022], and more up-to-date but less complete model structure can be found in the [Nimue documentation](https://web.archive.org/web/20221230232915/https://mrc-ide.github.io/nimue/index.html) [@winskillNimue]. Here we will describe the model in broad terms.

The baseline model is an age-stratified SEIRD model. In addition to classifying the population by infection status, the model also classifies people by vaccination status: they may be unvaccinated, vaccinated with 1 dose, vaccinated with 2 doses, vaccinated with 2 doses with waned protection, vaccinated 3 doses and vaccinated with 3 doses with waned protection. People are vaccinated according to vaccination dosage data from Our World in Data [@mathieuCoronavirusPandemicCOVID192020] together with a model of vaccine prioritisation. They progress to "waned" classes according to an assumed rate of waning of vaccine-derived immunity, which is itself time and dose-number-dependent. The model accounts for separate modes of vaccine action (infection blocking vs disease blocking), variant-dependent efficacy of vaccine and naturally derived immunity and age-dependent vaccination strategies. 

The model assumes homogenous mixing of the population. It models different kinds of infection -- "severe" and "non-severe" -- which have different age-dependent probabilities of progressing to recovery or death. Individuals who have recovered are assumed to be fully protected by natural immunity for an exponentially random duration duration which is, on average, around 250 days. Individuals who have been vaccinated have different levels of protection at different points in time depending on the dominant strain, and unless they receive additional doeses then protection is considered to have waned after an exponentially random period of time (on average, around 150 days). Vaccinated individuals are also considered to have a reduced likelihood of onward transmission, which also fades when the vaccine protection wanes.

The model features a mix of deterministic parameters and random parameters. We run an ensemble of simulations for each country, sampling the random parameters at the start of each run from pre-defined distributions. The initial number of cases seeding the pandemic in the country and the time series of reproduction numbers of infections in a fully susceptible population $R_t$ is then fit to the observed course of deaths during the time period associated with the simulation. 

The distributions of random parameters is given in Table \@ref(tab:trajectory-sampling-parameters). Parameters are sampled from each distribution independently except for the duration of vaccine-derived immunity, which is dependent on vaccine efficacy. 100 sets of parameters are sampled, and epidemic trajectories simulated for each. 95% intervals in our results refer to percentiles among the set of 100 sampled trajectories, so a 95% interval for deaths averted is the pair of numbers given by the 2.5th and the 97.5th percentile of deaths averted among all sampled trajectories. Note that the distributions given here are not exact representations of the models sampling distributions, which involve additional truncation steps.

The model of [@watsonoliverj.COVID19LMICReports2022] also samples probabilities of hospitalisation and mechanical ventilation for each trajectory, but these play no role in our analysis not already captured by the infection fatality rate, and so we haven't reported on the relevant parameters here. 

The shape parameters $\alpha_e^P, \beta_e^P$ that appear in Table \@ref(tab:trajectory-sampling-parameters) are the shape parameters associated with a beta distribution with mean $\mu_e^P$ and variance $0.005$. The mean $\mu_e^P$ is determined by estimating the efficacy of each vaccine platform $P$ against each variant (see Table \@ref(tab:vaccine-efficacy)). The mean vaccine durations are determined according to models of antibody decay, and are dependent on the sampled vaccine efficacies $\mu_e^P$ [@watsonoliverj.COVID19LMICReports2022].


```{r trajectory-sampling-parameters, tidy=FALSE, echo=FALSE, escape=FALSE, caption=c('Pandemic trajectory sampling distributions')}

sampling_dists <- data.frame(
    'Parameter' = c('Vaccine efficacy $V$', 'Vaccine duration $D$', 'Infection fatality rate with treatment $F$', 'Natural immunity duration $R$'), 
    'Distribution' = c(
      '$V\\sim \\mathrm{Beta}(\\alpha_{e}^P, \\beta_{e}^P), P = \\mathrm{vaccine\\; platform})$', 
      '$\\frac{1}{\\mathrm{Gamma}(\\alpha_{d}, \\beta_d)}$',
      '$F = \\mathrm{rescale}(X,f_{min}, f_{med}, f_{max})\\quad X\\sim \\mathrm{Beta}(2,2)$',
      '$R = \\mathrm{Gamma}(20, 4/73)$'
      ))

sampling_dists %>% kbl(
  booktabs = TRUE,
  linesep = "",
  caption = 'Distributions from which parameters are sampled. The parameters $\\alpha_e^P, \\beta_e^P$ are shape parameters for vaccine platform $P$ with associated efficacy $e$ (see text for additional details). The parameters $\\alpha_d, \\beta_d$ are shape and rate parameters for distribution of vaccine durations, determined by fitting an antibody decay curve to vaccine efficacies. $\\mathrm{rescale}(\\cdot, f_{min}, f_{med}, f_{max})$ is a function that maps $0$ to $f_{min}$, $0.5$ to $f_{med}$ and $1$ to $f_{max}$, linearly interpolating between each. $f_{min}, f_{med}, f_{max}$ are respectively the minimum, median, and maximum estimates of age-adjusted infection fatality.',
  centering = TRUE,
  escape = FALSE
)
```

```{r vaccine-platforms, tidy=FALSE, echo=FALSE, caption=c('Vaccine platforms for each country')}

sampling_dists <- data.frame(
    'Country' = c('USA', 'UK'), 
    'Platforms' = c(
      'mRNA, single dose, subunit', 'adenovirus, mRNA'))

sampling_dists %>% kbl(
  booktabs = TRUE,
  linesep = "",
  caption = 'Vaccine platforms used by each country.',
  centering = TRUE,
  escape = FALSE
)
```


```{r vaccine-efficacy, tidy=FALSE, echo=FALSE, caption=c('Model parameters - central estimates of vaccine efficacy')}

vaccine_efficacy <- read.csv('vaccine_efficacy_groups.csv')

vaccine_efficacy %>% kbl(
  booktabs = TRUE,
  linesep = "",
  col.names = c('Vaccine type', 'Doses', 'Strain', 'Protection against infection'),
  caption = 'Model parameters - central estimates of vaccine efficacy',
  centering = TRUE
)
```


We made one change to the model used by [@watsonoliverj.COVID19LMICReports2022]: by default, vaccination was assumed to reduce onward transmission by 50%, regardless of whether the dose was fresh or waned. We modified this to the schedule shown in Table \@ref(tab:onward). These represented an average between the estimates of [@eyreImpactSARSCoV2Vaccination2021a] for effectiveness at blocking onward transmission of Alpha and Delta and the estimate of [@tanInfectiousnessSARSCoV2Breakthrough2023] for effectiveness of the vaccine at blocking onward transmission of Omicron, as the period of the simulation included both Delta and Omicron waves.

```{r onward, tidy=FALSE, echo=FALSE, caption=c('Assumed relative transmissibility of infection by vaccinated individuals compared to unvaccinated')}
vaccine_onward_protection <- 
  data.frame(
    'Doses' = c(
      '1 dose (fresh)', 
      '2 doses (fresh)', 
      '2 doses (waned)',
      '3 doses (fresh)',
      '3 doses (waned stage 1)',
      '3 doses (waned stage 2)'
      ),
    'Protection from onward transmission' = c(
      '27%', 
      '27%', 
      '0%',
      '30%',
      '10%',
      '5%'
      )
    )

vaccine_onward_protection %>% kbl(
  booktabs = TRUE,
  col.names = c('Doses', 'Protection from onward transmission'),
  linesep = "",
  caption = 'Assumed relative transmissibility of infection by vaccinated individuals compared to unvaccinated',
  centering = TRUE
)
```


## Vaccine production

To produce counterfactual vaccine dosage timeseries, we suppose that vaccine approcal is brought forward by a number of days (we test approval coming 30, 60 and 90 day sooner). However, if approval is brought forward too much then manufacturers might have a harder time maintaining supply than they would have in the situation that actually played out. We model this by ... (see Tomas' entry in the shared doc)

Plots can currently be found in [Vaccine production model]

# Appendix C - Supplementary Results

## Impact of counterfactual vaccination scenarios

Illustrated in Figure \@ref(fig:cumulative-deaths).

```{r cumulative-deaths, fig.align="center", fig.cap=c("Cumulative deaths per scenario"), echo=FALSE}
cumulative_deaths_plot = ggplot(zoomed_deaths_averted,
                                aes(x = as.Date(date),
                                    y = cumulative_deaths_avg_perpop,
                                    color = counterfactual)) +
  geom_line() +
  geom_ribbon(
    aes(ymin = cumulative_deaths_025_perpop, ymax = cumulative_deaths_975_perpop, fill = counterfactual),
    alpha = 0.2,
    color = NA,
    show.legend = FALSE
  ) +
  geom_line(aes(y = baseline_cumulative_deaths_avg_perpop, colour = "baseline")) +
  facet_wrap( ~ country) +
  scale_colour_manual(
    aesthetics = c("fill", "colour"),
    labels = c('30 days', '60 days', '90 days', 'baseline'),
    values = c(
      counterfactual1_color,
      counterfactual2_color,
      counterfactual3_color,
      baseline_color
    )
  ) +
  labs(x = "Date", y = "Cumulative Deaths per 10 000 people", color = "Earlier vaccines by:")
  

plot(cumulative_deaths_plot)
```

# Appendix D - Sensitivity

## Baseline fit to reported deaths instead of excess deaths

Illustrated in Table \@ref(tab:reported-deaths-averted-table) and Figure \@ref(fig:reported-deaths-averted-plot).

```{r reported-deaths-averted-table, echo=FALSE}

differences_table <- deaths_averted_table_single(load_deaths_averted_table( excess = TRUE ))

differences_table %>%
  kbl(booktabs=TRUE, centering = TRUE, caption = "Averted deaths calculated on the basis of a model fit to excess deaths") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)

```

```{r reported-deaths-averted-plot, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario"), echo=FALSE}
deaths_averted = readRDS("data/deaths_averted_detail.Rds")

zoomed_deaths_averted = deaths_averted %>%
  filter(date > as.Date("2020-09-01") & date <= primary_last_date) %>%
  filter(counterfactual %in% cfacts_to_display) %>%
  mutate(counterfactual_label = mapvalues(
    counterfactual,
    from = cfacts_to_display,
    to = labels_to_display
  )) %>%
  mutate(counterfactual_label = factor(counterfactual_label,
                                       labels_to_display))

single_timeseries_plot = ggplot(zoomed_deaths_averted, aes(x = date)) +
  geom_line(aes(y = baseline_deaths_avg, colour = "baseline")) +
  geom_line(aes(y = deaths_avg, colour = "counterfactual")) +
  geom_ribbon(
    aes(ymin = deaths_025, ymax = deaths_975),
    alpha = 0.3,
    fill = colour_counterfactual
  ) +
  facet_grid(counterfactual_label ~ country,
             labeller = labeller(.rows=label_wrap_gen(width = 15))) +
  scale_colour_manual(
    labels = c('baseline', 'counterfactual'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  labs(x = "Date", y = "Daily Deaths", color = "Colour")
  

plot(single_timeseries_plot)


```

## Counterfactual scenario: what if US had doubled the rate of booster uptake?

Illustrated in Table \@ref(tab:deaths-averted-table-doubleboost).

```{r deaths-averted-table-doubleboost, echo=FALSE, results='asis'}
differences_table <- deaths_averted_table_single(
    load_deaths_averted_table(
      table_path = "data/doubleboost_deaths_averted_detail.Rds")
  )

differences_table %>% kbl(booktabs = TRUE, centering = TRUE, caption = "How would US outcomes change if boosters were adopted at twice the actual rate?") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)


```

## Counterfactual scenario: what if vaccine demand was lower?

Here, we examine the effect of lower vaccination demand. We consider vaccine demand to be 5% (Table \@ref(tab:deaths-averted-table-d95)) or 10% (Table \@ref(tab:deaths-averted-table-d90)) lower. We implemented this by asuming that daily vacinations were reduced by the respective amount. 

```{r deaths-averted-table-d95, echo=FALSE, results='asis'}
differences_table <- deaths_averted_table_single(
    load_deaths_averted_table(
      table_path = "data/d95_deaths_averted_detail.Rds")
  )

differences_table %>% kbl(booktabs = TRUE, centering = TRUE, caption = "How would outcomes change if vaccination demand were 5\\% lower?") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)


```

```{r deaths-averted-table-d90, echo=FALSE, results='asis'}
differences_table <- deaths_averted_table_single(
    load_deaths_averted_table(
      table_path = "data/d90_deaths_averted_detail.Rds")
  )

differences_table %>% kbl(booktabs = TRUE, centering = TRUE, caption = "How would outcomes change if vaccination demand were 10\\% lower") %>%
  pack_rows("United Kingdom to April 2021", 1, 3) %>%
  pack_rows("United Kingdom to July 2021", 4, 6) %>%
  pack_rows("United Kingdom to Jan 2022", 7, 9) %>%
  pack_rows("United States to April 2021", 10, 12) %>%
  pack_rows("United States to July 2021", 13, 15) %>%
  pack_rows("United States to Jan 2022", 16, 18) %>%
  kable_styling(font_size = 7)


```

## Sensitivity to parametric estimates

```{r deaths-averted-table-vei, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$veis %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to average vaccine efficacy against infection (VEI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)
```
```{r deaths-averted-table-durV, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$dur_Vs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of vaccine acquired immunity (DVI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)
```
```{r deaths-averted-table-durR, echo=FALSE, results='asis'}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$dur_Rs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of naturally acquired immunity (DNI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)

```
```{r deaths-averted-plot-vei, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to vaccine efficacy"), echo=FALSE}
deaths_averted_plots <- deaths_averted_plot('veis', percentiles, last_dates_to_display)
```
```{r deaths-averted-plot-durV, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to vaccine duration"), echo=FALSE}
deaths_averted_plots <- deaths_averted_plot('dur_Vs', percentiles, last_dates_to_display)
```
```{r deaths-averted-plot-durR, fig.align="center", fig.cap=c("DJ6 Daily deaths per scenario, sensitivity to natural immunity duration"), echo=FALSE}
deaths_averted_plots <- deaths_averted_plot('dur_Rs', percentiles, last_dates_to_display)
```
```{r vaccinated-plot-vei, fig.align="center", fig.cap=c("Effective vaccinated, sensitivity to vaccine duration"), echo=FALSE}
deaths_averted_plots <- vaccinated_plot('dur_Vs', percentiles, last_dates_to_display[3])
```
Illustrated in Table \@ref(tab:deaths-averted-table-vei), Table \@ref(tab:deaths-averted-table-durV), Table \@ref(tab:deaths-averted-table-durR), Figure \@ref(fig:deaths-averted-plot-vei), Figure \@ref(fig:deaths-averted-plot-durV), Figure \@ref(fig:deaths-averted-plot-durR), and Figure \@ref(fig:vaccinated-plot-vei).

# Vaccine production model

Figure \@ref(fig:prod-assumptions) shows the assumptions we made about the achievable production when vaccines were approved to be used 30, 60 and 90 days earlier.

```{r prod-assumptions, fig.align="center", fig.cap=c("DJ9 Vaccine production assumptions"), echo=FALSE}
detail_prod = readRDS("data/detail_production.Rds")
pfizer_prod = readRDS("data/pfizer_estimated_production.Rds")
moderna_prod = readRDS("data/moderna_estimated_production.Rds")

plot1 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = cumulative_available_vaccines)) +
  xlab("Date") +
  ylab("Cumulative vaccines available") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot2 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = daily_vaccines)) +
  xlab("Date") +
  ylab("Daily vaccines produced") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot3 = ggplot() +
  geom_line(data = moderna_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nModerna") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot4 = ggplot() +
  geom_line(data = pfizer_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nPfizer") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

combined_plot =
  ggarrange(plot3,
            plot4,
            plot2,
            plot1,
            nrow = 2,
            ncol = 2)
plot(combined_plot)
```


Figures \@ref(fig:prod-cap-uk) and \@rel{fig:prod-cap-usa} show our assumptions about the maximum number of vaccines that could be produced each day if vaccines were approved 30, 60 and 90 days earlier.

```{r prod-cap-uk, fig.align="center", fig.cap=c("DJ10 Limits from production UK"), echo=FALSE}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United Kingdom"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)
```


```{r prod-cap-usa, fig.align="center", fig.cap=c("DJ10 Limits from production UK"), echo=FALSE}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United States"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)

```

# Subsequent infection waves

While our main results present the number of deaths averted up to July 2021, we simulated the counterfactual scenarious up until the end of 2021. The longer timeline includes an additional wave of infections in both the US and the UK. Our model shows much more uncertainty over the effect of earlier vaccination timelines for this second wave of infections in the United States. In this country, for vaccines 90 days earlier, the 95% interval of simulation traces ranges from 36 739 to 221 093 extra lives saved by January 2022, and the intervals for 30 and 60 days earlier include some traces in which the earlier timelines resulted in a larger number of deaths in total. The UK, on the other hand, shows less variation in the impact of earlier timelines on the number of deaths in the second wave, and all counterfactuals result in fewer deaths at all points in time.

There are numerous ways that altering the timing of vaccinations can alter the pandemic trajectory for subsequent waves of infection. Effects that enter into our model are: 1. Earlier vaccine administration means that people will experience waning vaccine protection earlier, which could make them vulnerable if this waning coincides with a peak in infections (this is expected to increase infections at later dates) 2. Earlier vaccine administration can reduce the total number of infections immediately following the vaccination campaign, leading to fewer people having naturally acquired immunity at later dates (this is expected to increase infections at later dates) 3. Earlier vaccine administration can reduce the number of people infected at the start of a followup wave of infections (this is expected to increase the time taken to reach an infection peak and reduce the overall number of infections) 4. Earlier vaccine administration goes together with earlier administration of booster doses, so people may be more protected at followup waves due to having recently received a booster (this is expected to reduce the overall number of infections)

Earlier vaccine administration may also lead to behaviour change, which is not captured by our model. The effect of behaviour change downstream of vaccination is also hard to predict. If earlier vaccination leads to fewer infections in deaths at a given point in time, and as a result people relax social distancing behaviour, then our model might overestimate short-term infections and deaths in the early vaccination scenario (though people would of course be benefiting from reduced social distancing). At the same time, such a situation would lead to a higher rate of natural immunity, which could reduce the impact of a subsequent infection wave.

We hypothesise that differences between the US and the UK that lead to the divergent results may be: - Booster doses were much more popular in the UK than in the US. On Jan 1 2022 50.5% of the UK population had received booster doses compared to 23.4% of the US population (Mathieu et al., 2020) - Our model predicts that in the short run, earlier vaccinations suppress the level of infections in the UK more than they do in the US.

To test the hypothesis that differential booster doses explain some of the difference, we ran a simulation where the rate of booster doses in the US was counterfactually adjusted so that the coverage of booster doses in both the US and UK were equal on Jan 1 2022 (to achieve this, we rougly doubled the booster dose coverage in the US). The results are in Supplementary Table 8. We note that this counterfactual doesn’t drastically reduce the uncertainty over long run deaths averted in the US, suggesting that the differences between the US and the UK estimates are driven by something other than differences in booster adoption.

While we show that earlier vaccinations have large benefits to mortality in infections waves occurring close to the vaccination campaign, and our best guess is that earlier vaccinations yield large benefits to mortality overall, we are substantially less certain of the latter conclusion. Due to uncertainty about the effects of early vaccination on followup waves, we are substantially uncertain about the overall size and sign of the mortality benefit for COVID-19. Furthermore, future pandemics (or even followup waves of COVID-19) are going to differ from the period we studied, and the overall benefits of accelerated vaccinations are likely to be sensitive to these differences.

[^21]: [https://www.npr.org/sections/health-shots/2020/12/17/947628608/how-will-moderna-meet-the-demand-for-its-covid-19-vaccine](https://www.npr.org/sections/health-shots/2020/12/17/947628608/how-will-moderna-meet-the-demand-for-its-covid-19-vaccine)

[^22]: [https://www.pfizer.com/news/press-release/press-release-detail/pfizer-and-biontech-celebrate-historic-first-authorization](https://www.pfizer.com/news/press-release/press-release-detail/pfizer-and-biontech-celebrate-historic-first-authorization)

[^23]: [https://www.npr.org/sections/health-shots/2020/12/17/947628608/how-will-moderna-meet-the-demand-for-its-covid-19-vaccine](https://www.npr.org/sections/health-shots/2020/12/17/947628608/how-will-moderna-meet-the-demand-for-its-covid-19-vaccine)

[^24]: This is an estimate based on them producing 2.5 times more than Moderna in 2020 but getting faster approval and  running at full speed longer.

[^25]: [https://www.pfizer.com/news/press-release/press-release-detail/pfizer-and-biontech-co-develop-potential-covid-19-vaccine](https://www.pfizer.com/news/press-release/press-release-detail/pfizer-and-biontech-co-develop-potential-covid-19-vaccine)

[^26]: [https://time.com/5807669/coronavirus-vaccine-moderna/](https://time.com/5807669/coronavirus-vaccine-moderna/),[ https://s29.q4cdn.com/435878511/files/doc_events/2020/03/30/Moderna-COVID-19-Impact-on-Trials-Transcript-(03.30.20).pdf](https://s29.q4cdn.com/435878511/files/doc_events/2020/03/30/Moderna-COVID-19-Impact-on-Trials-Transcript-(03.30.20).pdf)

[^27]: [https://www.pfizer.com/news/articles/shot_of_a_lifetime_how_two_pfizer_manufacturing_plants_upscaled_to_produce_the_covid_19_vaccine_in_record_time](https://www.pfizer.com/news/articles/shot_of_a_lifetime_how_two_pfizer_manufacturing_plants_upscaled_to_produce_the_covid_19_vaccine_in_record_time)

[^28]: [https://investors.modernatx.com/news/news-details/2020/Moderna-and-Catalent-Announce-Collaboration-for-Fill-Finish-Manufacturing-of-Modernas-COVID-19-Vaccine-Candidate/default.aspx](https://investors.modernatx.com/news/news-details/2020/Moderna-and-Catalent-Announce-Collaboration-for-Fill-Finish-Manufacturing-of-Modernas-COVID-19-Vaccine-Candidate/default.aspx)

[^29]: [https://www.fda.gov/news-events/press-announcements/fda-takes-key-action-fight-against-covid-19-issuing-emergency-use-authorization-first-covid-19](https://www.fda.gov/news-events/press-announcements/fda-takes-key-action-fight-against-covid-19-issuing-emergency-use-authorization-first-covid-19)

[^30]: [https://www.fda.gov/news-events/press-announcements/fda-takes-additional-action-fight-against-covid-19-issuing-emergency-use-authorization-second-covid](https://www.fda.gov/news-events/press-announcements/fda-takes-additional-action-fight-against-covid-19-issuing-emergency-use-authorization-second-covid)

[^31]: They also got earlier approval in the UK.
