BIBLIOGRAPHY

---
output:
  pdf_document: default
  html_document: default
---

\newpage

# (APPENDIX) Supplementary material {-}

\beginsupplement

# Appendicies {-}

# Manufacturing constraints

If vaccines were approved earlier, production capacity would probably have been a major constraint on the number of doses delivered at the beginning of the mass vaccination campaigns. As mentioned in the main text, hypothesising the extent to which production is a binding constraint is complicated by the fact that new manufacturing capacity (or expansion of existing capacity) was put in place in anticipation of regulatory approval. Therefore it is likely that an earlier approval would have led the vaccine makers to expand production faster. In our base case model we ignore this, making our assumption on available doses a conservative one. 

While we model the United States and the United Kingdom using separate epidemiological models, we do not differentiate between countries when it comes to manufacturing capacity, reasoning that if a country was able to approve a vaccine earlier it would jump to the front of the vaccine deliveries queue. We also ignore AstraZeneca vaccine production since the production of Pfizer and Moderna vaccines are enough to satisfy demand in the UK in the initial phase of vaccinations. (As a conservative assumption, we still assume that vaccine effectiveness in the UK is a mix of mRNA vaccines and the one manufactured by AstraZeneca.)

To estimate the potential number of vaccines that would have been available throughout 2020, we used information about the actual manufacturing timelines. We assume that in 2020 Moderna produced 20 million doses and Pfizer/BioNTech produced 50 million [@NPR2020How, @Pfizer2020Pfizer]. Additionally, we know Moderna was producing 500,000 doses a day at the time their vaccine was authorised and assume Pfizer/BioNTech was producing 1 million doses a day at the time they received authorization [@NPR2020How][^21]. These numbers, combined with estimates of scale-up dates, give us a possible function of daily vaccine production over 2020.

We assume a linear scale-up of production from March to July, followed by constant production until two weeks before authorization by the FDA. Specifically, we assume Pfizer/BioNTech started working on scale up from their announced partnership on March 17th and Moderna from March 23rd [@Pfizer2020BioNTech, @Time2020As, @Moderna2020Edited]. The start of mass production in July (for calculations we use July 31st) was assumed based on reports that Pfizer might have been ready to mass produce by then, with Moderna entering a partnership with Catalent in June to do their fill-and-finish processing with an estimated start at the beginning of Q3 [@PfizerNDShot, @Moderna2020Moderna]. Time of vaccine authorization was December 11th Pfizer/BioNTech, and December 18th for Moderna meaning another scale up in manufacturing started on November 27th for Pfizer and December 4th for Moderna [@FDA2020FDA, @FDA2020Second].

At this point, we assume companies already knew that they would most likely get authorization[^22] and were able to do another linear scale-up to a production level they would maintain until the end of the year. After the end of 2020 we assume production is unconstrained, i.e. it is as if infinite vaccines were available each day since 01-01-2021. This assumption is not important since even with a 90-day shift of vaccination series the production constraint stops being binding before the end of 2020. Additionally, we assume that if a country did approve vaccines earlier, they would nearly have all manufactured vaccines to themself. 

In the case of the United Kingdom this means that in all scenarios of earlier vaccination the production constraint does not bind at all, since the amount of vaccines is very large compared with their population size. In the case of the United States the production only binds in the case when the vaccination is shifted by 90 days.

# Vaccine hesitancy

A useful framework for vaccine hesitancy is provided by the “5C’s” model, which considers the following drivers: confidence in vaccine safety, complacency, convenience, risk calculation, and sense of collective responsibility [@Machingaidze2021Understanding]. For the purposes of our model, confidence in vaccine safety is the factor most clearly affected by a hypothetical earlier approval. We used surveys evaluating intention to be vaccinated and surveys evaluating confidence in health authorities to approximate the effect of vaccine hesitancy within our timeframe of interest. 

In the US and UK, the rate of respondents who intended to be vaccinated when vaccines were approved was 41% and 73% respectively. These proportions at 90 days prior to approval were 39% in the US and 71% in the UK, with the lowest rates of intention to be vaccinated at 36% in October 2020 in the US and 61% in November 2020 in the UK [@YouGov2023COVID, @Freeman2022COVID]. In a separate series of surveys in the US, 85% of respondents had a lot or a fair amount of confidence in health authorities at time of vaccine approval compared to 89% in April 2020 [@Kennedy2022Americans]. In the UK, surveys asking respondents their confidence in the country’s health system indicated 73% were confident at time of approval and 77% were confident 90 days prior [@YouGov2023Confidence]. By these estimations, vaccine hesitancy and confidence in health authorities hovered around similar levels during our period of interest. 

@Dzieciolowska2021COVID conducted a study of 2,761 health care workers in Canada in December 2020 to assess vaccine hesitancy, with 19.1% of participants indicating they were not willing to receive the COVID vaccine. Among participants who refused the vaccine, 82% indicated the novelty of the vaccine was very or somewhat important in their decision and 60% indicated that they did not have enough time to make a decision. An approval of vaccines using accelerated timelines could obviously negatively impact hesitancy rooted in vaccine novelty and feelings of needing more time to decide. However, the main hypothetical we discuss in our paper is accelerating efficacy testing and starting safety testing early. This also suggests that while the impact may be non-negligible, it may not be large.

@Joshi2021Predictors found that the global average vaccine acceptance rate was 86% in March 2020, dropped to 54% in July 2020, and increased to 72% in September 2020.  By these estimates, if vaccines were approved during summer of 2020, hesitancy may have appreciably impacted uptake. However, it appears that vaccine hesitancy would likely not have impacted uptake if vaccines had been available up to 90 days earlier in the UK and US, with surveys evaluating intention to be vaccinated and trust in health authorities varying by no more than a few percentage points during this time. It is possible these trends would have been impacted by a concern among the general public for vaccines approved quickly, but these effects are difficult to quantify.

\newpage

# Epidemiological methods

To model alternative vaccination timelines, we took an existing compartmental model that simulated the COVID-19 pandemic through 2020 and 2021 including the impact of vaccinations (the "baseline model") and re-ran the simulation freezing all parameters except the vaccination timelines (the "counterfactual models").

The baseline pandemic model used the model behind [@watsonoliverj.COVID19LMICReports2022]. This model is implemented in open source R packages Squire [@hoganWithincountryAgebasedPrioritisation2021, @walkerImpactCOVID19Strategies2020, @watsonLeveragingCommunityMortality2021] and SirCOVID [@baguelinSircovidSIRModel2022]. An older version of this model is described in detail in the supplementary material of @walkerImpactCOVID19Strategies2020, and details of many parameters -- again, for an older version of the model -- can be found in the supplementary material of @watsonGlobalImpactFirst2022. Additional incomplete information about the parametrisation of the version of the model we used can be found at [COVID-19 LMIC Reports documentation](https://web.archive.org/web/20221230232942/https://mrc-ide.github.io/global-lmic-reports/parameters.html) [@watsonoliverj.COVID19LMICReports2022], and more up-to-date but less complete model structure can be found in the [Nimue documentation](https://web.archive.org/web/20221230232915/https://mrc-ide.github.io/nimue/index.html) [@winskillNimue]. Here we will describe the model in broad terms.

The baseline model is an age stratified SEIRD model. In addition to classifying the population by infection status, the model also classifies people by vaccination status: they may be unvaccinated, vaccinated with 1 dose, vaccinated with 2 doses, vaccinated with 2 doses with waned protection, vaccinated 3 doses and vaccinated with 3 doses with waned protection. People are vaccinated according to vaccination dosage data from Our World in Data [@mathieuCoronavirusPandemicCOVID192020] together with a model of vaccine prioritisation. They progress to "waned" classes according to an assumed rate of waning of vaccine-derived immunity, which is itself time and dose-number-dependent. The model accounts for separate modes of vaccine action (infection blocking vs disease blocking), variant-dependent efficacy of vaccine and naturally derived immunity and age-dependent vaccination strategies. 

The model assumes homogenous mixing of the population. It models different kinds of infection -- "severe" and "non-severe" -- which have different age-dependent probabilities of progressing to recovery or death. Individuals who have recovered are assumed to be fully protected by natural immunity for an exponentially random duration which is, on average, around 250 days. Individuals who have been vaccinated have different levels of protection at different points in time depending on the dominant strain, and unless they receive additional doses then protection is considered to have waned after an exponentially random period of time (on average, around 150 days). Vaccinated individuals are also considered to have a reduced likelihood of onward transmission, which also fades when the vaccine protection wanes.

The model features a mix of deterministic parameters and random parameters. We run an ensemble of simulations for each country, sampling the random parameters at the start of each run from pre-defined distributions. The initial number of cases seeding the pandemic in the country and the time series of reproduction numbers of infections in a fully susceptible population $R_t$ is then fit to the observed course of deaths during the time period associated with the simulation. 

```{r trajectory-sampling-parameters, tidy=FALSE, echo=FALSE, escape=FALSE, caption=c('Pandemic trajectory sampling distributions'), out.height="40%"}

sampling_dists <- data.frame(
    'Parameter' = c('Vaccine efficacy $V$', 'Vaccine duration $D$', 'Infection fatality rate with treatment $F$', 'Natural immunity duration $R$'), 
    'Distribution' = c(
      '$V\\sim \\mathrm{Beta}(\\alpha_{e}^P, \\beta_{e}^P), P = \\mathrm{vaccine\\; platform})$', 
      '$\\frac{1}{\\mathrm{Gamma}(\\alpha_{d}, \\beta_d)}$',
      '$F = \\mathrm{rescale}(X,f_{min}, f_{med}, f_{max})\\quad X\\sim \\mathrm{Beta}(2,2)$',
      '$R = \\mathrm{Gamma}(20, 4/73)$'
      ))

sampling_dists %>% kbl(
  booktabs = TRUE,
  linesep = "",
  caption = 'Distributions from which parameters are sampled. The parameters $\\alpha_e^P, \\beta_e^P$ are shape parameters for vaccine platform $P$ with associated efficacy $e$ (see text for additional details). The parameters $\\alpha_d, \\beta_d$ are shape and rate parameters for distribution of vaccine durations, determined by fitting an antibody decay curve to vaccine efficacies. $\\mathrm{rescale}(\\cdot, f_{min}, f_{med}, f_{max})$ is a function that maps $0$ to $f_{min}$, $0.5$ to $f_{med}$ and $1$ to $f_{max}$, linearly interpolating between each. $f_{min}, f_{med}, f_{max}$ are respectively the minimum, median, and maximum estimates of age-adjusted infection fatality.',
  centering = TRUE,
  escape = FALSE
)
```
The distributions of random parameters is given in Table \@ref(tab:trajectory-sampling-parameters). Parameters are sampled from each distribution independently except for the duration of vaccine-derived immunity, which is dependent on vaccine efficacy. 100 sets of parameters are sampled, and epidemic trajectories simulated for each. 95% intervals in our results refer to percentiles among the set of 100 sampled trajectories, so a 95% interval for deaths averted is the pair of numbers given by the 2.5th and the 97.5th percentile of deaths averted among all sampled trajectories. Note that the distributions given here are not exact representations of the models sampling distributions, which involve additional truncation steps.

The model of @watsonoliverj.COVID19LMICReports2022 also samples probabilities of hospitalisation and mechanical ventilation for each trajectory, but these play no role in our analysis not already captured by the infection fatality rate, and so we haven't reported on the relevant parameters here. 

```{r vaccine-efficacy, tidy=FALSE, echo=FALSE, caption=c('Model parameters - central estimates of vaccine efficacy'), out.height="40%"}

vaccine_efficacy <- read.csv('vaccine_efficacy_groups.csv')

vaccine_efficacy %>% kbl(
  booktabs = TRUE,
  linesep = "",
  col.names = c('Vaccine type', 'Doses', 'Strain', 'Protection against infection'),
  caption = 'Model parameters - central estimates of vaccine efficacy',
  centering = TRUE
)
```
The shape parameters $\alpha_e^P, \beta_e^P$ that appear in Table \@ref(tab:trajectory-sampling-parameters) are the shape parameters associated with a beta distribution with mean $\mu_e^P$ and variance $0.005$. The mean $\mu_e^P$ is determined by estimating the efficacy of each vaccine platform $P$ against each variant (see Table \@ref(tab:vaccine-efficacy)). The mean vaccine durations are determined according to models of antibody decay, and are dependent on the sampled vaccine efficacies $\mu_e^P$ [@watsonoliverj.COVID19LMICReports2022].

```{r onward, tidy=FALSE, echo=FALSE, caption=c('Assumed relative transmissibility of infection by vaccinated individuals compared to unvaccinated'), out.height="40%"}
vaccine_onward_protection <- 
  data.frame(
    'Doses' = c(
      '1 dose (fresh)', 
      '2 doses (fresh)', 
      '2 doses (waned)',
      '3 doses (fresh)',
      '3 doses (waned stage 1)',
      '3 doses (waned stage 2)'
      ),
    'Protection from onward transmission' = c(
      '27%', 
      '27%', 
      '0%',
      '30%',
      '10%',
      '5%'
      )
    )

vaccine_onward_protection %>% kbl(
  booktabs = TRUE,
  col.names = c('Doses', 'Protection from onward transmission'),
  linesep = "",
  caption = 'Assumed relative transmissibility of infection by vaccinated individuals compared to unvaccinated',
  centering = TRUE
)
```
We made one change to the model used by [@watsonoliverj.COVID19LMICReports2022]: by default, vaccination was assumed to reduce onward transmission by 50%, regardless of whether the dose was fresh or waned. We modified this to the schedule shown in Table \@ref(tab:onward). These represented an average between the estimates of [@eyreImpactSARSCoV2Vaccination2021a] for effectiveness at blocking onward transmission of Alpha and Delta and the estimate of [@tanInfectiousnessSARSCoV2Breakthrough2023] for effectiveness of the vaccine at blocking onward transmission of Omicron, as the period of the simulation included both Delta and Omicron waves.

\newpage

# Sensitivity To Model Parameters

We explored various alternative model configurations in order to assess the sensitivity of our conclusion to modelling assumptions. Compared to a hypothetical ideal model that yields correct counterfactual assessments, the model we employ is likely to differ in a number of ways:

- It may differ structurally. We might anticipate some structural differences - for example, overdispersion in the distribution of contact rates is not captured by our model, but overdispersion in this distribution was typically found to be high [@endoEstimatingOverdispersionCOVID192020]. However, there may be other structural differences that we do not anticipate
- The assumed distributions of input parameters may differ in our model and in the hypothetical ideal

Both of these differences mean that the fitted values of $R_t$ are also likely to differ between our model and the hypothetical ideal, and hence they may yield substantially different assessments of counterfactual scenarios. If it turns out that counterfactual assessments are very sensitive to input parameter values, then we might conclude that our estimates are likely to differ substantially from the "true" counterfactual. Because of the prospect of structural differences, this is true even if we have done a very good job of estimating parameters. If our results are robust to variation of parameters within a reasonable range, then if we believe that our model is capable of yielding a good approximation to the ideal for some "reasonable" parameter choices, we should also think our counterfactual assessments are a good approximation to ideal counterfactual assessments.

We assess the impact of three different parameter estimates on our overall results. To do this, we examine the average conclusion from the model runs featuring the top and bottom deciles of each of the following parameters:

- The average infection-blocking efficacy of one and two doses of the vaccine
- The average duration for protection due to one and two doses of the vaccine
- The average duration of protection due to natural immunity

In both countries, the estimates of deaths averted showed little sensitivity to the estimated duration of natural immunity (Table \@ref(tab:deaths-averted-table-durR). The estimates of deaths averted in the UK was *also* relatively insensitive to the estimated vaccine efficacy, though the estimate in the US was more sensitive to this value (it is worth noting that in the case of the US the model explored captured a wider range of variation in vaccine efficacy) - see Table \@ref(tab:deaths-averted-table-vei). Note that for long-run estimates of deaths averted in the US (up to Jan 2022), higher estimates of vaccine efficacy were associated with much greater uncertainty over the number of deaths averted. This may be due to the fact that, given a more effective vaccine, waning immunity will have a larger impact on the end results.

The estimate of long-run deaths averted in the US was extremely sensitive to the estimated duration of vaccine derived immunity, with a difference of 14 days in this parameter estimate yielding long-run estimates of deaths averted that ranged from 32,255 to *negative* 11,456 (that is, 11,456 extra deaths) for a 30-day advance in the vaccination schedule. Notably, short estimates of vaccine duration were also associated with extreme decreases in the uncertainty over the number of deaths averted. Note that the short run estimates of deaths averted (up to July 2021) were robustly positive, but were also substantially more uncertain for longer estimates of the duration of vaccine protection.

Our model is already very uncertain about the effect of earlier vaccinations on the second infection wave in the US. However, the high sensitivity of this figure to the estimated duration of vaccine-derived protection offers an extra reason to be unsure that the model is providing us with an accurate assessment of the counterfactual impacts on this timescale.

We also run an identical analysis to our main analysis, except with a model fit to reported numbers of COVID-19 deaths instead of COVID-19 deaths estimated from excess mortality. The results are reported in Table \@ref(tab:reported-deaths-averted-table) and Figure \@ref(fig:reported-deaths-averted-plot). This method yields larger estimates of deaths averted than our main method, particularly up to July 2021 where the estimates are close to the 95th percentile estimate for the main method. This is in spite of the fact that the total estimated number of deaths under this method is somewhat lower than under the excess mortality method.

```{r deaths-averted-table-vei, echo=FALSE, results='asis', out.height="35%"}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$veis %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to average vaccine efficacy against infection (VEI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)
```
```{r deaths-averted-table-durV, echo=FALSE, results='asis', out.height="35%"}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$dur_Vs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of vaccine acquired immunity (DVI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)
```
```{r deaths-averted-table-durR, echo=FALSE, results='asis', out.height="35%"}
differences_tables <- deaths_averted_table_multiple(join_counterfactual_tables(parameter_variants, percentiles, last_dates_to_display))
differences_tables$dur_Rs %>% kbl(booktabs = TRUE, centering = TRUE, caption = "Sensitivity to duration of naturally acquired immunity (DNI)") %>%
  pack_rows("UK to April 2021", 1, 6) %>%
  pack_rows("UK to July 2021", 7, 12) %>%
  pack_rows("UK to Jan 2022", 13, 18) %>%
  pack_rows("US to April 2021", 19, 24) %>%
  pack_rows("US to July 2021", 25, 30) %>%
  pack_rows("US to Jan 2022", 31, 36) %>%
  kable_styling(font_size = 7)

```
```{r deaths-averted-plot-vei, fig.align="center", fig.cap=c("Daily averted deaths under counterfactual scenarios, with sensitivity to vaccine efficacy. Shaded area indicates 95\\% confidence interval."), echo=FALSE, out.height="35%"}
deaths_averted_plot_veis <- deaths_averted_plot('veis', percentiles, last_dates_to_display)
plot(deaths_averted_plot_veis)
```
```{r deaths-averted-plot-durV, fig.align="center", fig.cap=c("Daily averted deaths under counterfactual scenarios, with sensitivity to vaccine duration. Shaded area indicates 95\\% confidence interval."), echo=FALSE, out.height="35%"}
deaths_averted_plot_dur_Vs <- deaths_averted_plot('dur_Vs', percentiles, last_dates_to_display)
plot(deaths_averted_plot_dur_Vs)
```
```{r deaths-averted-plot-durR, fig.align="center", fig.cap=c("Daily averted deaths under counterfactual scenarios, with sensitivity to natural immunity duration. Shaded area indicates 95\\% confidence interval."), echo=FALSE, out.height="35%"}
deaths_averted_plot_dur_Rs <- deaths_averted_plot('dur_Rs', percentiles, last_dates_to_display)
plot(deaths_averted_plot_dur_Rs)
```
```{r vaccinated-plot-vei, fig.align="center", fig.cap=c("Effective vaccinated, sensitivity to vaccine duration"), echo=FALSE, out.height="35%"}
effective_vaccinated_plot_dur_Vs <- vaccinated_plot('dur_Vs', percentiles, last_dates_to_display[3])
```

\newpage
# Vaccine production model

```{r prod-assumptions, fig.align="center", fig.cap=c("Vaccine production assumptions"), echo=FALSE, out.height="40%"}
detail_prod = readRDS("data/detail_production.Rds")
pfizer_prod = readRDS("data/pfizer_estimated_production.Rds")
moderna_prod = readRDS("data/moderna_estimated_production.Rds")

plot1 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = cumulative_available_vaccines)) +
  xlab("Date") +
  ylab("Cumulative vaccines available") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot2 = ggplot() +
  geom_line(data = detail_prod,
            aes(x = date, y = daily_vaccines)) +
  xlab("Date") +
  ylab("Daily vaccines produced") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot3 = ggplot() +
  geom_line(data = moderna_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nModerna") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

plot4 = ggplot() +
  geom_line(data = pfizer_prod,
            aes(x = date, y = daily_production)) +
  xlab("Date") +
  ylab("Daily vaccines produced\nPfizer") +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6))

combined_plot =
  ggarrange(plot3,
            plot4,
            plot2,
            plot1,
            nrow = 2,
            ncol = 2)
plot(combined_plot)
```
Figure \@ref(fig:prod-assumptions) shows the assumptions we made about the achievable production when vaccines were approved to be used 30, 60 and 90 days earlier.

\newpage
```{r prod-cap-uk, fig.align="center", fig.cap=c("Cumulative vaccine production and distribution in the U.K. under counterfactual and baseline scenarios"), echo=FALSE, out.height="40%"}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United Kingdom"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)
```
```{r prod-cap-usa, fig.align="center", fig.cap=c("Cumulative vaccine production and distribution in the U.S. under counterfactual and baseline scenarios"), echo=FALSE, out.height="40%"}
production = readRDS("data/counterfactual_production.Rds")

country_name = "United States"

country_cfact = counterfactuals %>%
  filter(country == country_name) %>%
  mutate(shifted_by = factor(
    shifted_by,
    levels = c("baseline", "30", "60", "90"),
    labels = c(
      "Baseline",
      "Vaccines 30 days sooner",
      "Vaccines 60 days sooner",
      "Vaccines 90 days sooner"
    )
  ))
  country_prod = production %>% filter(country == country_name)
  

dj10 = ggplot() +
  geom_line(data = country_prod,
            aes(
              x = as.Date(date),
              y = cumulative_available_vaccines,
              color = "Production"
            )) +
  geom_line(data = country_cfact,
            aes(
              x = as.Date(date),
              y = total_vacc,
              color = "Vaccination"
            )) +
  labs(x = "Date",
       y = "Cumulative production / vaccinations",
       color = "Colour") +
  ggtitle(country_name) +
  scale_colour_manual(
    labels = c('Production', 'Vaccination'),
    values = c(colour_baseline, colour_counterfactual)
  ) +
  scale_y_continuous(labels = unit_format(unit = "M", scale = 1e-6)) +
  facet_wrap(~ shifted_by)

plot(dj10)

```
Figure \@ref(fig:prod-cap-uk) and \@ref(fig:prod-cap-usa) show our assumptions about the maximum number of vaccines that could be produced each day if vaccines were approved 30, 60 and 90 days earlier.

[^21]: This is an estimate based on them producing 2.5 times more than Moderna in 2020 but getting faster approval and running at full speed longer.

[^22]: They also got earlier approval in the UK.
